---
title: '2. Virtual species defined by response to environmental variables'
output:
  html_document:
    fig_caption: yes
    keep_md: yes
    mathjax: "http://cdn.mathjax.org/mathjax/latest/MathJax.js?config=TeX-AMS-MML_HTMLorMML"
    theme: yeti
    toc: true
    toc_float: true
---

The first approach to generate a virtual species consists in defining its 
response functions to each of the environmental variables contained in our 
`RasterStack`. These responses are then combined to calculcate the environmental
suitability of the virtual species.

The function providing this approach is `generateSpFromFun`.

## 2.1. An introduction example

Before we start using the package, let's prepare our first simulation of
virtual species.

We want to generate a virtual species with two environmental variables, 
the annual mean temperature `bio1` and annual mean precipitation `bio2`.
We want to generate a tropical species, i.e., living in hot and humid 
environments. We can define bell-shaped response functions to these two 
variables, as in the following figure:

```{r vspload, echo=FALSE, fig.width=8, fig.height = 4, fig.cap = "Fig. 2.1 Example of bell-shaped response functions to bio1 and bio2, suitable for a tropical species."}
library(virtualspecies)
library(geodata)
worldclim <- worldclim_global(var = "bio", res = 10,
                              path = tempdir())
names(worldclim) <- paste0("bio", 1:19)

a <- formatFunctions(x = worldclim[[c(1, 12)]], bio1 = c(fun = "dnorm", 
                                                         mean = 25, sd = 5),
                     bio12 = c(fun = "dnorm", mean = 4000, sd = 2000),
                     rescale = FALSE)
rm(a)

```

These bell-shaped functions are in fact gaussian distributions functions of the form:

$$ f(x, \mu, \sigma) = \frac{1}{\sigma\sqrt{2\pi}}\, e^{-\frac{(x - \mu)^2}{2 \sigma^2}}$$

If we take the example of bio1 above, the mean $\mu$ is equal to 25&deg;C and
standard deviation $\sigma$ is equal to 5&deg;C. Hence the 
response function for bio1 is:

$$ f(bio1) = \frac{1}{5\sqrt{2\pi}}\, e^{-\frac{(bio1 - 25)^2}{2 \times 5^2}} $$

In R, it is very simple to obtain the result of the equation above, with the
function `dnorm`: 

```{r resp2, cache = TRUE}
# Suitability of the environment for bio1 = 15 degree C
dnorm(x = 15, mean = 25, sd = 5)
```

The idea with `virtualspecies` is to keep the same simplicity when generating 
a species: we will use the `dnorm` function to generate our virtual species.  

Let's start with the package now. We load environmental data first:

```{r resp2.1, cache = TRUE}
library(geodata)
worldclim <- worldclim_global(var = "bio", res = 10,
                              path = tempdir())
# Rename bioclimatic variables for simplicity
names(worldclim) <- paste0("bio", 1:19)
```


The first step is to provide to the helper function `formatFunctions` which
responses you want for which variables:

```{r resp3, cache = TRUE}
library(virtualspecies)

my.parameters <- formatFunctions(bio1 = c(fun = 'dnorm', mean = 25, sd = 5),
                                 bio12 = c(fun = 'dnorm', mean = 4000, 
                                           sd = 2000))
```

After that, the generation of the virtual species is fairly easy:
```{r resp4, cache = TRUE, fig.cap = "Fig. 2.2 Environmental suitability of the generated virtual species"}

# Generation of the virtual species
my.first.species <- generateSpFromFun(raster.stack = worldclim[[c("bio1", "bio12")]],
                                      parameters = my.parameters,
                                      plot = TRUE)
my.first.species
```

Congratulations! You have just generated your first virtual species 
distribution, with the default settings. In the next section you will learn 
about the simple, yet important settings of this function.

## 2.2. Customisation of parameters

The function `generateSpFromFun` proceeds into four important steps:

1. The response to each environmental variable is calculated according to the
user-chosen functions.
2. Each response is rescaled between 0 and 1. This step can be disabled.
3. The responses are combined together to compute the environmental suitability.
The user can choose how to combine the responses.
4. The environmental suitability is rescaled between 0 and 1. This step can be 
disabled.

### 2.2.1. Response functions
You can use any existing function with `generateSpFromFun`, as long as you 
provide the necessary parameters. 
For example, you can use the function `dnorm` shown above, if you provide the
parameters `mean` and `sd`:
`formatFunctions(bio1 = c(fun = 'dnorm', mean = 25, sd = 5))`. Naturally you can 
change the values of `mean` and `sd` to your needs.

You can also use the basic functions provided with the package:

* Linear function: `formatFunctions(bio1 = c(fun = 'linearFun', a = 1, b = 0))`
$$ f(bio1) = a * bio1 + b $$
```{r resp4.1, echo = FALSE, cache = TRUE, fig.cap = "Fig. 2.3 Linear response function", fig.width = 3, fig.height = 2.5, results = "hide", message=FALSE, fig.align = 'center'}
par(mar = c(3.1, 3.1, 0.1, 0.1))
formatFunctions(bio1 = c(fun = 'linearFun', a = 1, b = 0),
                x = worldclim[["bio1"]])
```

* Quadratic function: `formatFunctions(bio1 = c(fun = 'quadraticFun', a = -1, b = 2, c = 0))`
$$ f(bio1) = a \times bio1^2 + b \times bio1 + c$$
```{r resp4.2, echo = FALSE, cache = TRUE, fig.cap = "Fig. 2.4 Quadratic response function", fig.width = 3, fig.height = 2.5, results = "hide", message=FALSE, fig.align = 'center'}
par(mar = c(3.1, 3.1, 0.1, 0.1))
formatFunctions(bio1 = c(fun = 'quadraticFun', a = -1, b = 2, c = 0),
                x = worldclim[["bio1"]])
```

* Logistic function: `formatFunctions(bio1 = c(fun = 'logisticFun', beta = 150, alpha = -5))`
$$ f(bio1) = \frac{1}{1 + e^{\frac{bio1 - \beta}{\alpha}}} $$
```{r resp4.3, echo = FALSE, cache = TRUE, fig.cap = "Fig. 2.5 Logistic response function", fig.width = 3, fig.height = 2.5, results = "hide", message=FALSE, fig.align = 'center'}
par(mar = c(3.1, 3.1, 0.1, 0.1))
formatFunctions(bio1 = c(fun = 'logisticFun', beta = 15, alpha = -5),
                x = worldclim[["bio1"]])
```
* Normal function defined by extremes: 
`formatFunctions(bio1 = c(fun = 'custnorm', mean = 25, diff = 5, prob = 0.99))`
This function allows you to set extremum of a normal curve. In the example 
above, we define a response where the optimum is 25&deg;C (`mean = 25`), and
99% of the area under the curve (`prob = 0.99`) will be comprised between 20
and 30&deg;C (`diff = 50`).
```{r resp4.4, echo = FALSE, cache = TRUE, fig.cap = "Fig. 2.6 Normal function defined by extremes", fig.width = 3, fig.height = 2.5, message=FALSE, results = "hide", fig.align = 'center'}
par(mar = c(3.1, 3.1, 0.1, 0.1))
formatFunctions(bio1 = c(fun = 'custnorm', mean = 25, diff = 5, prob = 0.99), 
                x = worldclim[["bio1"]])
```

* Beta response function (Oksanen & Minchin, 2002,
_Ecological Modelling_ __157__:119-129): 
`formatFunctions(bio1 = c(fun = 'betaFun', p1 = 0, p2 = 25, alpha = 0.9, gamma = 0.08))`
$$ f(bio1) = (bio1 - p1)^{\alpha} * (p2 - bio1)^{\gamma} $$
```{r resp4.5, fig.cap = "Fig. 2.7 Beta response function",  echo = FALSE, cache = TRUE, fig.width = 3, fig.height = 2.5, results = "hide", message=FALSE, fig.align = 'center'}
par(mar = c(3.1, 3.1, 0.1, 0.1))
formatFunctions(bio1 = c(fun = 'betaFun', p1 = 0, p2 = 250, alpha = 0.9,
                         gamma = 0.08),
                x = worldclim[["bio1"]])
```

* Or you can create your own functions (see the section
[How to create your own response functions](#how-to-create-and-use-your-own-response-functions) 
if you need help for this).

### 2.2.2. Rescaling of individual response functions

This rescaling is performed because if you use very different response function 
for different variables, (_e.g._, a Gaussian distribution function with a linear
function), then the responses may be disproportionate among variables. By 
default this rescaling is enabled (`rescale.each.response = TRUE`), but it can
be disabled (`rescale.each.response = FALSE`).

### 2.2.3. Combining response functions

There are three main possibilities to combine response functions to calculate 
the environmental suitability, as defined by the parameters `species.type` and
`formula`:

* `species.type = "additive"`: the response functions are added.
* `species.type = "multiplicative"`: the response functions are multiplied.
This is the default behaviour of the function.
* `formula = "bio1 + 2 * bio2 + bio3"`: if you choose a formula, then the 
response functions are combined according to your formula (parameter
`species.type` is then ignored).

For example, if you want to generate a species with the same partial responses 
as in 2.1, but with a strong importance for temperature, then you can specify
the formula : `formula = "2 * bio1 + bio12"`

```{r resp5, cache = TRUE, fig.cap = "Fig. 2.3 Environmental suitability of the generated virtual species"}
library(virtualspecies)

my.parameters <- formatFunctions(bio1 = c(fun = 'dnorm', mean = 25, sd = 5),
                                 bio12 = c(fun = 'dnorm', mean = 4000, sd = 2000))

# Generation of the virtual species
new.species <- generateSpFromFun(raster.stack = worldclim[[c("bio1", "bio12")]],
                                 parameters = my.parameters,
                                 formula = "2 * bio1 + bio12",
                                 plot = TRUE)
new.species
```

One can even make complex interactions between partial responses:

```{r resp6, cache = TRUE, fig.cap = "Fig. 2.3 Environmental suitability of the generated virtual species"}
library(virtualspecies)

my.parameters <- formatFunctions(bio1 = c(fun = 'dnorm', mean = 25, sd = 5),
                                 bio12 = c(fun = 'dnorm', mean = 4000, sd = 2000))

# Generation of the virtual species
new.species <- generateSpFromFun(raster.stack = worldclim[[c("bio1", "bio12")]],
                                 parameters = my.parameters,
                                 formula = "3.1 * bio1^2 - 1.4 * sqrt(bio12) * bio1",
                                 plot = TRUE)
new.species
```

Note that this is an example to show the possibilities of the function; I have
no idea of the relevance of such a relationship!

### 2.2.4. Rescaling of the final environmental suitability

This final rescaling is performed because the combination of the different 
responses can lead to very different range of values. It is therefore necessary
to allow environmental suitabilities to be comparable among virtual species,
and should not be disabled unless you have very precise reasons to do it. The 
argument `rescale` controls this rescaling (`TRUE` by default).

## 2.3. How to create and use your own response functions

An important aspect of `generateSpFromFun` is that you can create and use your 
own response functions. In this section we will see how we can do that in
practice.  
We will take the example of a simple linear function:
$$ f(x, a, b) = ax + b$$

Our first step will be to create the function in R:

```{r custfun1, cache = TRUE}
linear.function <- function(x, a, b)
{
  a*x + b
}
```

That's it! You created a function called `linear.function` in R.

Let's try it now:
```{r custfun2, cache = TRUE}
linear.function(x = 0.5, a = 2, b = 1)

linear.function(x = 3, a = 4, b = 1)

linear.function(x = -20, a = 2, b = 0)
```

It seems to work properly. Now we will use `linear.function` to generate a
virtual species distribution. We want a species responding linearly to the 
annual mean temperature, and with a gaussian to the annual precipitations:

```{r custfun3, cache = TRUE, fig.cap = "Fig 2.3 Environmental suitability of the generated virtual species"}
my.responses <- formatFunctions(bio1 = c(fun = "linear.function", a = 1, b = 0),
                                bio12 = c(fun = "dnorm", mean = 1000, sd = 1000))

generateSpFromFun(raster.stack = worldclim[[c("bio1", "bio12")]],
                  parameters = my.responses, plot = TRUE)
```

And it worked! Your turn now!

