---
title: '5. Generating random species'
output:
  html_document:
    fig_caption: yes
    keep_md: yes
    mathjax: "http://cdn.mathjax.org/mathjax/latest/MathJax.js?config=TeX-AMS-MML_HTMLorMML"
    theme: yeti
    toc: true
    toc_float: true
---

The `virtualspecies` package embeds a function to randomly generate virtual 
species `generateRandomSp`, with many customisable options to constrain the 
randomisation process.

Let's take a look:

```{r rand1, eval = T, cache = TRUE, fig.height = 6, fig.width = 10, fig.cap = "Fig. 5.1 A species randomly generated with `generateRandomSp`"}
library(virtualspecies)
library(geodata)
worldclim <- worldclim_global(var = "bio", res = 10,
                              path = tempdir())
names(worldclim) <- paste0("bio", 1:19)

my.stack <- worldclim[[c("bio2", "bio5", "bio6", "bio12", "bio13", "bio14")]]
random.sp <- generateRandomSp(my.stack)
random.sp

```

We can see that the species was generated using a PCA approach. Indeed,
[as explained in the PCA section](#second-approach-generate-virtual-species-with-a-principal-components-analysis), when you have a lot of variables, it becomes very difficult to 
generate a species with realistic environmental requirements. Hence, by default 
the function `generateRandomSp` uses a PCA approach if you have 6 or more 
variables, and a 'response functions' approach if you have less than 6 
variables.

In the next sections I detail the many possible customisations for the function
`generateRandomSp`.

## 5.1. General parameters

### 5.1.1. Choosing the approach

You can choose `approach = `:

* `"automatic"`: a 'response' approach is used if you have less than 6 variables 
and a 'PCA' approach is used if you have 6 or more variables
* `"random"`: a random approach is chosen (response or PCA)
* `"response"`: to use a response approach
* `"pca"`: to use a pca approach

### 5.1.2. Rescaling of the environmental suitability

By default, `rescale = TRUE`, which means that the environmental suitability is
rescaled between 0 and 1.

### 5.1.3. Conversion to presence-absence

You can choose to enable the conversion of environmental suitability to 
presence-absence. To do this, set `convert.to.PA = TRUE`. You can customise
the conversion:

* choose the conversion method with `PA.method`. You should leave it to
`probability` unless you have a very particular reason to use the `threshold` 
approach.
* define the parameters `alpha`, `beta` and `species.prevalence` exactly as
explained in the convert to presence-absence section, or leave them as they
are to create a random conversion into presence-absence

## 5.2. Parameters specific to a 'response' approach

### 5.2.1. Define the possible response functions

At the moment, four relations are possible for a random generation of virtual 
species: Gaussian (`gaussian`), linear (`linear`), logistic (`logistic`) and 
quadratic (`quadratic`) relations. By default, all the relation types are used.
You can choose to use any combination with the argument `relations`:

```{r rand2, eval = T, cache = TRUE, fig.cap = "Fig. 5.2 A species randomly generated with `generateRandomSp`, with gaussian and logistic response functions"}
# A species with gaussian and logistic response functions
random.sp1 <- generateRandomSp(worldclim[[1:3]], 
                              relations = c("gaussian", "logistic"),
                              convert.to.PA = FALSE)

random.sp1
```

```{r rand3, fig.width = 5, fig.height = 5, eval = T, cache = TRUE, fig.cap = "Response functions of the randomly generated species"}
plotResponse(random.sp1)
```


```{r rand4, eval = T, cache = TRUE, fig.cap = "Fig. 5.4 A species randomly generated with `generateRandomSp`, with gaussian response functions only"}
# A purely gaussian species
random.sp2 <- generateRandomSp(worldclim[[1:3]], 
                              relations = "gaussian",
                              convert.to.PA = FALSE)

random.sp2
```

```{r rand5, fig.width = 5, fig.height = 5, eval = T, cache = TRUE, fig.cap = "Fig. 5.5 Response functions of the randomly generated species"}
plotResponse(random.sp2)
```


### 5.2.2. Rescale individual response functions

As explained in the section on the 'response' approach, you can choose to 
rescale or not each individual response function, with the argument 
`rescale.each.response`. `TRUE` by default.

### 5.2.3. Try to find a realistic species

An important issue with the generation of random responses to the environment,
is that you can obtain a species willing to live in summer temperatures of 
35&deg;C and winter temperature of -50&deg;C. This may interesting for 
generating species from another planet, but you are probably more interested
in generating species that can actually live on Earth. There is therefore an
option to do that, activated by default: `realistic.sp`.

When activating this argument, the function will proceed step-by-step to try 
defining a realistic species. At step one, one of the variable is chosen, and 
the program randomly determines a response function for this variable. Then, it
will compute the environmental suitability of this species. At step two, the 
program will pick a second variable, and will constrain its random generation 
depending on the environmental suitability obtained at step one. For example,
if at step 2 a gaussian response is picked, then the mean will be chosen in 
areas where the species had a high environmental suitability. Then, the 
environmental suitability is recalculated on the basis of the first two response
functions. At step 3, another variable is picked, a response function randomly
generated with respect to areas where the species already has a high 
suitability, and so on until there are no variables left. While this process 
can help, it does not always work, and can provide completely unrealistic 
results also. In this case, you should try different runs, or switch to a
'PCA' approach.

Let's see an example :

```{r rand6, eval = T, cache = TRUE, fig.cap = c("Fig. 5.6 A species randomly generated, constrained to find realistic environmental requirements", "Fig. 5.7 A species randomly generated, with no constraints")}
realistic.sp <- generateRandomSp(worldclim[[c(1, 5)]],
                                 realistic.sp = TRUE,
                                 convert.to.PA = FALSE) 
 
 

unrealistic.sp <- generateRandomSp(worldclim[[c(1, 5)]],
                                   realistic.sp = FALSE,
                                   convert.to.PA = FALSE)
```

Note that you can always let the function randomly determine the conversion to
presence-absence by changing the argument `convert.to.PA = TRUE` (be careful:
since parameters are defined randomly, the results can be bizarre), or you can 
do it by yourself later:

```{r rand7, fig.width = 10, fig.height = 6, eval = T, cache = TRUE, fig.cap = "Fig. 5.8 Modification of the conversion threshold of the previously generated species in fig. 5.6", cache = TRUE}
realistic.sp <- convertToPA(realistic.sp, beta = 0.5) 
```

If, in spite of all your attempts, you are struggling to obtain satisfactory 
species, then perhaps you should try the 'PCA' approach.

### 5.2.4. Define how response functions are combined to compute the
environmental suitability

You can define how response functions are combined with the argument `sp.type`, 
exactly as explained in the 'response' approach section:

* `species.type = "additive"`: the response functions are added.
* `species.type = "multiplicative"`: the response functions are multiplied 
(default value). 
* `species.type = "mixed"`: there is a mix of additions and products to 
calculate the environmental suitability, randomly generated.

## 5.3. Parameters specific to a 'PCA' approach

### 5.3.1. Generate a random species with a PCA approach on a low-memory
computer

As explained in the PCA section, if you have low-memory computer, or are 
working with very large raster files (large extent, fine resolution), then 
you can still perform the PCA by using a sample of the pixels of your raster.

To do this, set `sample.points = TRUE`, and choose the number of pixels to 
sample, depending on your computer's memory, with `nb.points`.

```{r rand8, fig.width = 10, fig.height = 6, eval = T, cache = TRUE, fig.cap = "Fig 5.9 A virtual species randomly generated from a PCA approach, with a memory-safe procedure"}
# A safe run for a low memory computer
safe.run.sp <- generateRandomSp(worldclim[[c(1, 5, 6)]], 
                                sample.points = TRUE,
                                nb.points = 1000)
```

### 5.3.2. Generate stenotopic or eurytopic species

You can modify the `niche.breadth` argument to generate stenotopic
(`niche.breadth = 'narrow'`), eurytopic (`niche.breadth = 'wide'`), or 
any type of species (`niche.breadth = 'any'`).
